---
title: "Fitting Exercise"
format: html
---

## Loading and Checking the data
The raw dataset contains repeated time‐series measurements of drug concentration (DV) over time (TIME) for each individual (ID). The data also includes dosing information (DOSE, AMT) and basic demographics (AGE, SEX, RACE WT, HT).

```{r}
library(tidyverse)
library(here)


dat_raw <- read_csv(here("fitting-exercise","Mavoglurant_A2121_nmpk.csv"))

dim(dat_raw)
head(dat_raw)
```


## Quick plot of concentration over time

```{r}
ggplot(dat_raw, aes(x = TIME, y = DV, group = ID)) +
  geom_line(alpha = 0.5) +
  facet_wrap(~ DOSE) +
  labs(
    x = "Time",
    y = "DV (drug concentration)"
  )
```


Plotting DV versus TIME (grouped by ID and stratified by DOSE): concentrations are highest shortly after dosing and decline over time. Higher dose groups appear to have higher concentration curves overall.

## Remove OCC = 2

```{r}
dat1 <- dat_raw %>%
  filter(OCC == 1)

dim(dat1)
table(dat_raw$OCC)

```

Some individuals have multiple occasions (OCC=1 and OCC=2), indicating repeated dosing/measurement periods. To keep the exercise simple and avoid having multiple trajectories per person, I kept only OCC=1 and removed OCC=2.

## Compute total drug amount per individual (Y)

```{r}
y_df <- dat1 %>%
  filter(TIME != 0) %>%
  group_by(ID) %>%
  summarize(Y = sum(DV, na.rm = TRUE))

dim(y_df)
head(y_df)

```



## Extract baseline rows (TIME = 0)

```{r}
base_df <- dat1 %>%
  filter(TIME == 0)

dim(base_df)
head(base_df)

```


## Join baseline data with total drug (Y)

```{r}
dat_person <- base_df %>%
  left_join(y_df, by = "ID")

dim(dat_person)
head(dat_person)

```


## Final cleaning

```{r}
dat_clean <- dat_person %>%
  mutate(
    SEX  = factor(SEX),
    RACE = factor(RACE)
  ) %>%
  select(Y, DOSE, AGE, SEX, RACE, WT, HT) %>%
  mutate(
    SEX  = forcats::fct_drop(SEX),
    RACE = forcats::fct_drop(RACE)
  )

dim(dat_clean)
head(dat_clean)

```


## Summary statistics

```{r}
summary(dat_clean)

dat_clean %>% count(DOSE)
dat_clean %>% count(SEX)
dat_clean %>% count(RACE)


```


## Relationship between Y and key predictors

```{r}
# Y vs DOSE 
ggplot(dat_clean, aes(x = factor(DOSE), y = Y)) +
  geom_boxplot() +
  labs(x = "DOSE", y = "Total drug (Y)", title = "Y by DOSE")

# Y vs SEX
ggplot(dat_clean, aes(x = SEX, y = Y)) +
  geom_boxplot() +
  labs(x = "SEX", y = "Total drug (Y)", title = "Y by SEX")

# Y vs AGE 
ggplot(dat_clean, aes(x = AGE, y = Y)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "AGE", y = "Total drug (Y)", title = "Y vs AGE")
```

```{r}
ggplot(dat_clean, aes(x = WT, y = Y)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "WT", y = "Total drug (Y)", title = "Y vs Weight")
```


```{r}
ggplot(dat_clean, aes(x = HT, y = Y)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "HT", y = "Total drug (Y)", title = "Y vs Height")
```

## Distributions

```{r}
dat_clean %>%
  pivot_longer(cols = c(Y, AGE, WT, HT),
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distributions of numeric variables")
```

## Pair/correlation plots

```{r}
# numeric-only correlation matrix
cor_mat <- dat_clean %>%
  select(Y, DOSE, AGE, WT, HT) %>%
  cor()

round(cor_mat, 2)
```

This suggests a strong positive relationship between Y and DOSE (cor ≈ 0.72), which is also visible in the boxplot of Y by dose. AGE shows little relationship with Y in this dataset. WT and HT are moderately correlated (cor ≈ 0.60).


### Linear regression: Y - DOSE

```{r}
library(tidymodels)

set.seed(123)

# split into training/testing
split1 <- initial_split(dat_clean, prop = 0.8)
train1 <- training(split1)
test1  <- testing(split1)

# model specification
lm_spec <- linear_reg() %>%
  set_engine("lm")

# workflow
wf_lm_dose <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE)

# fit
fit_lm_dose <- fit(wf_lm_dose, data = train1)

# predict on test + metrics
pred_lm_dose <- predict(fit_lm_dose, new_data = test1) %>%
  bind_cols(test1 %>% select(Y))

metrics(pred_lm_dose, truth = Y, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq"))
```

The linear model using DOSE as the only predictor explains approximately 69% of the variability in Y (R² ≈ 0.69). The RMSE is about 483 units, meaning that predictions are on average about 483 units away from the observed total drug value.



### Linear regression: Y - all predictors

```{r}
# workflow with all predictors
wf_lm_all <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE + AGE + SEX + RACE + WT + HT)

# fit
fit_lm_all <- fit(wf_lm_all, data = train1)

# predict on test + metrics
pred_lm_all <- predict(fit_lm_all, new_data = test1) %>%
  bind_cols(test1 %>% select(Y))

metrics(pred_lm_all, truth = Y, estimate = .pred) %>%
  filter(.metric %in% c("rmse", "rsq"))
```

Adding AGE, SEX, RACE, WT, and HT did not improve predictive performance. In fact, the test-set RMSE increased slightly (483 to 494) and R² decreased (0.687 to 0.662). This suggests that DOSE already captures most of the explainable variation in Y, and the additional predictors add little signal (or add noise) in this dataset.

### Logistic regression: SEX - DOSE

```{r}
set.seed(123)

split2 <- initial_split(dat_clean, prop = 0.8, strata = SEX)
train2 <- training(split2)
test2  <- testing(split2)

# logistic model specification
log_spec <- logistic_reg() %>%
  set_engine("glm")


wf_log_dose <- workflow() %>%
  add_model(log_spec) %>%
  add_formula(SEX ~ DOSE)

# fit
fit_log_dose <- fit(wf_log_dose, data = train2)

# predict probabilities
pred_log_dose <- predict(fit_log_dose, test2, type = "prob") %>%
  bind_cols(predict(fit_log_dose, test2)) %>%
  bind_cols(test2 %>% select(SEX))

# compute metrics
metric_set(accuracy, roc_auc)(pred_log_dose,
                              truth = SEX,
                              estimate = .pred_class,
                              .pred_1)

```

The model shows high accuracy (0.84), but the ROC-AUC is ~0.49, which is essentially random (≈0.5). The high accuracy is likely driven by class imbalance (most observations are in one SEX category), so a model that predicts the majority class most of the time can achieve high accuracy without being truly predictive. Overall, DOSE does not appear to meaningfully predict SEX.

### Logistic regression: SEX - all predictors


```{r}

set.seed(123)

# split
split2 <- initial_split(dat_clean, prop = 0.8, strata = SEX)
train2 <- training(split2)
test2  <- testing(split2)

# logistic model specification
log_spec <- logistic_reg() %>%
  set_engine("glm")

# with all predictors
wf_log_all <- workflow() %>%
  add_model(log_spec) %>%
  add_formula(SEX ~ DOSE + AGE + RACE + WT + HT)

# fit
fit_log_all <- fit(wf_log_all, data = train2)

# predict probabilities + classes on test set
pred_log_all <- predict(fit_log_all, test2, type = "prob") %>%
  bind_cols(predict(fit_log_all, test2)) %>%   # adds .pred_class
  bind_cols(test2 %>% select(SEX))

# compute metrics
metric_set(accuracy, roc_auc)(
  pred_log_all,
  truth = SEX,
  estimate = .pred_class,
  .pred_1
)

```

The model achieved very high performance on this test split, but the test set is small and imbalanced, so these metrics may be unstable or overly optimistic. Results should be interpreted carefully.

### Summary 

Overall, DOSE was a strong predictor of the continuous outcome Y, explaining a substantial portion of its variability. Adding additional baseline predictors did not improve performance meaningfully. For the categorical outcome SEX, predictive performance varied depending on the model and evaluation strategy, and results should be interpreted carefully due to class imbalance. This exercise illustrates the complete modeling workflow, including data splitting, model fitting, prediction, and evaluation.

# Module 10: Model Improvement

## Data Prep


```{r}
# remove RACE variable
data6 <- dat_person %>%
  mutate(
    SEX = factor(SEX)
  ) %>%
  select(Y, DOSE, AGE, SEX, WT, HT) %>%
  mutate(
    SEX = forcats::fct_drop(SEX)
  )

# set seed
rngseed <- 1234
set.seed(rngseed)

# data split
data_split <- initial_split(data6, prop = 3/4)

# create train/test sets
train_data <- training(data_split)
test_data  <- testing(data_split)
```

## Model Fitting


```{r}
library(tidymodels)

# model specification
lm_mod <- linear_reg() %>%
  set_engine("lm")

# fit model 1 (Y ~ DOSE)
lm_fit_10_1 <- lm_mod %>%
  fit(Y ~ DOSE, data = train_data)

lm_fit_10_1
```


```{r}
# fit model 2 (Y ~ all predictors)
lm_fit_10_2 <- lm_mod %>%
  fit(Y ~ ., data = train_data)

lm_fit_10_2
```

## Model Performance Assessment 1


```{r}
# predictions for model 1
preds_10_1 <- predict(lm_fit_10_1, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))

rmse(preds_10_1, truth = Y, estimate = .pred)
rsq(preds_10_1, truth = Y, estimate = .pred)
```

```{r}
# predictions for model 2
preds_10_2 <- predict(lm_fit_10_2, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))

rmse(preds_10_2, truth = Y, estimate = .pred)
rsq(preds_10_2, truth = Y, estimate = .pred)
```


```{r}
# null model
null_mod <- null_model(mode = "regression") %>%
  set_engine("parsnip")

null_fit <- null_mod %>%
  fit(Y ~ 1, data = train_data)

null_preds <- predict(null_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))

rmse(null_preds, truth = Y, estimate = .pred)
```

The null model shows the highest RMSE, indicating the poorest performance. The DOSE-only model performs better, while the model including all predictors achieves the lowest RMSE. This suggests that adding AGE, SEX, WT, and HT improves prediction of Y compared to using DOSE alone.

## Model Performance Assessment 2

```{r}
# create 10-fold cross-validation splits
set.seed(rngseed)
folds <- vfold_cv(train_data, v = 10)
```

```{r}
# perform cross-validation for model 1
cv_fit1 <- lm_mod %>%
  fit_resamples(
    Y ~ DOSE,
    resamples = folds,
    metrics = metric_set(rmse)
  )

collect_metrics(cv_fit1)
```


```{r}
# perform cross-validation for model 2
cv_fit2 <- lm_mod %>%
  fit_resamples(
    Y ~ .,
    resamples = folds,
    metrics = metric_set(rmse)
  )

collect_metrics(cv_fit2)
```

Using 10-fold cross-validation, the RMSE for the DOSE-only model was approximately 691, while the model including all predictors achieved a lower RMSE of about 646. Compared to the earlier training-based estimates, the RMSE values are slightly higher, which is expected since cross-validation provides a more realistic estimate of performance on unseen data.

The model with all predictors still performs better than the DOSE-only model, indicating that the additional variables provide some predictive value when evaluated more robustly. The standard errors indicate some variability across folds, reflecting the limited sample size and the impact of random data splitting.

```{r}
# run CV again with a different seed
rngseed2 <- 5678
set.seed(rngseed2)

folds2 <- vfold_cv(train_data, v = 10)

cv_fit3 <- lm_mod %>%
  fit_resamples(
    Y ~ DOSE,
    resamples = folds2,
    metrics = metric_set(rmse)
  )

collect_metrics(cv_fit3)
cv_fit4 <- lm_mod %>%
  fit_resamples(
    Y ~ .,
    resamples = folds2,
    metrics = metric_set(rmse)
  )

collect_metrics(cv_fit4)

```

Repeating the cross-validation with a different random seed resulted in slightly different RMSE values for both models. This variation is expected due to randomness in the data splitting process. However, the overall pattern remained consistent, with the model including all predictors continuing to outperform the DOSE-only model.

# **This section is added by Austin Hill**

```{r}
# load packages 
library(tidyverse)
library(tidymodels)
```

```{r}
# Create a combined data frame of observed and predicted values for all 3 original models

# Model 1 plotting data
model1_plot_df <- preds_10_1 %>%
  rename(observed = Y) %>%
  mutate(model = "Model 1: Y ~ DOSE")

# Model 2 plotting data
model2_plot_df <- preds_10_2 %>%
  rename(observed = Y) %>%
  mutate(model = "Model 2: Y ~ all predictors")

# Null model plotting data
null_plot_df <- null_preds %>%
  rename(observed = Y) %>%
  mutate(model = "Null model")

# Combine all three models into one data frame
all_model_preds <- bind_rows(
  model1_plot_df,
  model2_plot_df,
  null_plot_df
)

# Inspect combined plotting data
glimpse(all_model_preds)
```

```{r}
# Plot observed vs predicted for all 3 models

ggplot(
  all_model_preds,
  aes(x = observed, y = .pred, color = model, shape = model)
) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Observed vs Predicted Values for Three Models",
    x = "Observed Values",
    y = "Predicted Values",
    color = "Model",
    shape = "Model"
  ) +
  theme_minimal()
```

```{r}
# Create residuals for Model 2

model2_resid_df <- preds_10_2 %>%
  rename(observed = Y) %>%
  mutate(residual = .pred - observed) %>%
  select(observed, .pred, residual)

glimpse(model2_resid_df)
```
```{r}
# Create symmetric y-axis limits

max_abs_resid <- max(abs(model2_resid_df$residual), na.rm = TRUE)
```
```{r}
# Plot predicted vs residuals for Model 2

ggplot(model2_resid_df, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_cartesian(ylim = c(-max_abs_resid, max_abs_resid)) +
  labs(
    title = "Predicted Values vs Residuals for Model 2",
    x = "Predicted Values",
    y = "Residuals (Predicted - Observed)"
  ) +
  theme_minimal()
```

```{r}
# Re-set seed and create 100 bootstrap samples using the training data

set.seed(rngseed)

boot_splits <- bootstraps(train_data, times = 100)

boot_splits

```

```{r}
# Fit Model 2 to each bootstrap sample and predict on the original training data

boot_pred_list <- map(boot_splits$splits, function(split) {
  
  boot_train <- analysis(split)
  
  # Re-fit Model 2 using the same model spec as before
  boot_fit <- lm_mod %>%
    fit(Y ~ ., data = boot_train)
  
  # Predict on the original training data
  predict(boot_fit, new_data = train_data) %>%
    pull(.pred)
})
```
```{r}
# Convert bootstrap predictions to a matrix
# Rows = training observations
# Columns = bootstrap samples

boot_pred_mat <- do.call(cbind, boot_pred_list)

dim(boot_pred_mat)
```

```{r}
# Create summary data frame with observed values, original point estimates, and bootstrap summaries

boot_summary_df <- tibble(
  observed = train_data$Y,
  point_estimate = preds_10_2$.pred,
  boot_mean = apply(boot_pred_mat, 1, mean, na.rm = TRUE),
  boot_median = apply(boot_pred_mat, 1, median, na.rm = TRUE),
  lower_ci = apply(boot_pred_mat, 1, quantile, probs = 0.025, na.rm = TRUE),
  upper_ci = apply(boot_pred_mat, 1, quantile, probs = 0.975, na.rm = TRUE)
)

glimpse(boot_summary_df)
```

```{r}
# Plot observed values vs original predictions,bootstrap median, and bootstrap CI limits

ggplot(boot_summary_df, aes(x = observed)) +
  geom_point(aes(y = point_estimate, color = "Original Prediction"), alpha = 0.6) +
  geom_point(aes(y = boot_median, color = "Bootstrap Median"), alpha = 0.6) +
  geom_point(aes(y = lower_ci, color = "Lower Confidence Interval"), alpha = 0.5) +
  geom_point(aes(y = upper_ci, color = "Upper Confidence Interval"), alpha = 0.5) +
  
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  
  scale_color_manual(
    name = "Prediction Type",
    values = c(
      "Original Prediction" = "black", 
      "Bootstrap Median" = "blue", 
      "Lower Confidence Interval" = "red", # Added missing comma here
      "Upper Confidence Interval" = "green" # Fixed capitalization to match above
    )
  ) +
  
  coord_equal(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    title = "Observed vs Predicted Values with Bootstrap Uncertainty",
    x = "Observed Values",
    y = "Predicted Values"
  ) +
  theme_minimal()
```
# Discussion
This plot shows that the model is moderately accurate, as the points generally cluster around the dashed 45-degree line, though there is a noticeable amount of scatter. The vertical distance between the green (upper) and red (lower) points reveals maybe some prediction uncertainty. The black and blue dots are very tightly clustered. This means the model's central tendency is reliable and isn't being wildly tossed around by small changes in the training data.